Verification of Stencil Computations with Spatial Specifications

### 2016-11-11 paper.tex
giving stencil definition earlier

\begin{definition}[Base induction variable]
  An integer variable is a \emph{base induction variable} if is the
  control variable of a ``for'' loop (or equivalent, \eg{},
  \texttt{do} in Fortran for our implementation) incremented by $1$
  per iteration. The variable is interpreted as a based induction
variable only within the scope of the loop body. For brevity we
say \emph{induction variable}, which are ranged over by $i, j, k$
throughout.
%TODO: consider mentioning derived, non-based induction variables
% as we did before.
\end{definition}

\begin{definition}[Array subscripts and indices]
  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression
  which reads from an array $a$ at an \emph{index} specified by a
  comma-separated sequence of integer expressions denoted $\bar{e}$ or 
  in expanded form as $(e_1, \ldots, e_n)$. An
  \emph{index} $\bar{e}$ is called \emph{relative} if it is defined in
  terms of induction variables.
\end{definition}

\begin{definition}[Neighbourhood index]
\label{def:neighbour}
  For an array subscript $a(\bar{e})$ an index $\bar{e}$ is called a
  \emph{neighbourhood index} if every $e \in \bar{e}$ is of the form
  $e \equiv i$ or $e \equiv i + c$ or $e \equiv i - c$ where $c$ is an
  integer constant. That is, a neighbourhood index is a relative index
  where each component is a constant translation of a base induction
  variable. (The relation $\equiv$ here identifies terms up-to
  commutativity of $+$ and the inverse relationship of $+$ and $-$
  \eg{}, $(-b) + i \equiv i - b$).  %We classify neighbourhood indices
%  by the predicate $\neigh_I$ where $I$ is a set of induction
%  variables. We omit $I$ when clear from the context.
\end{definition}

\dnote{Define: Stencil kernel, or maybe stencil assignment,
 and stencil much earlier than previously?}

Specification declarations associate (via
the operator $\texttt{::}$) a specification to one or more array
variables $\bar{v}$.% specifying the shape of indexing expressions for
%arrays $\bar{v}$ that are simultaenously permitted and expected within
%the code to which the specification is given.

\begin{definition}[Consistency, informal]
  A stencil specification defines a \emph{region} which, informally,
  is an n-dimensional rectangle which is possibly infinite in some
  dimensions. A stencil kernel in code is \emph{consistent} with a
  specification on array variable $a$ if every array index on $a$
\end{definition}

@@

A \term{forward} region of depth $n$ specifies continguous
neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
inclusively for some induction variable $i$.  Similarly, a
\term{backward} region of depth $n$ specifies contiguous neighbour
indices in dimension $d$ from $i$ to $i - n$, and \term{centered}
of depth $n$ from $i - n$ to $i + n$. A \term{pointwise}
stencil specifies a neighbour index $i$. For example the
following shows three specifications with three consistent stencil
kernels reading from arrays \term{a}, \term{b}, and \term{c} respectively:
%%
\begin{minted}{fortran}
!= stencil forward(depth=2, dim=1) :: a
d(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
!= stencil backward(depth=2, dim=1) :: b
d(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
!= stencil centered(depth=1, dim=2) :: c
d(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
\end{minted}
%%
Note that every dimension need not be specified for; in the above, the
specifications on lines $1$ and $5$ are the principal specifications
for code on lines $2$ and $6$, but the specification on line
$3$ is not principal for line $4$;
we return to the notion of principality later.

The \term{forward}, \term{backward}, and \term{centered} regions may
all have an additional attribute \term{nonpointwise} which marks that
the region does not include indices which are purely base induction
variables, \ie{}, offset by $0$.  For example, the following is a
\term{nonpointwise} \term{backward} stencil
%
\begin{minted}{fortran}
!= stencil backward(depth=2, dim=1, nonpointwise) :: a
b(i) = a(i-1) + a(i-2)
\end{minted}
%
The \term{nonpointwise} attribute cannot be used on \term{pointwise}
regions.

>>>>

\begin{definition}[Base induction variable]
  An integer variable is a \emph{base induction variable} (or simply
\emph{induction variable} for brevity) if it is the
  control variable of a ``for'' loop (or equivalent, \eg{},
  \texttt{do} in Fortran for our implementation) incremented by $1$
  per iteration. The variable is interpreted as a based induction
variable only within the scope of the loop body. Induction varaibles 
are ranged over by $i, j, k$ throughout.
%TODO: consider mentioning derived, non-based induction variables
% as we did before.
\end{definition}

\begin{definition}[Array subscripts and indices]
  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression
  which reads from an array $a$ at an \emph{index} specified by a
  comma-separated sequence of integer expressions denoted $\bar{e}$ or
  in expanded form as $(e_1, \ldots, e_n)$. An \emph{index} $\bar{e}$
  is called \emph{relative} if each $e \in \bar{e}$ is defined in
  terms of an induction variable.

  Relative indices contrast with absolute indices where each $e \in
  \bar{e}$ is a constant integer expression relative to the containing
  loop, \ie{}, $e$ is not defined in terms of an induction variable.
\end{definition}

\begin{definition}[Neighbourhood index]
\label{def:neighbour}
  For an array subscript $a(\bar{e})$ an index $\bar{e}$ is called a
  \emph{neighbourhood index} if every $e \in \bar{e}$ is of the form
  $e \equiv i$ or $e \equiv i + c$ or $e \equiv i - c$ where $c$ is an
  integer constant. That is, a neighbourhood index is a relative index
  where each component is a constant translation of a base induction
  variable. (The relation $\equiv$ here identifies terms up-to
  commutativity of $+$ and the inverse relationship of $+$ and $-$
  \eg{}, $(-b) + i \equiv i - b$). 
\end{definition}

\begin{definition}[Stencil computations]
  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
  collection of arrays of arbitrary (possible differing)
  dimensionalities.  
  A \emph{stencil computation} comprises an iteration by a set $I$ of at
  most $n$ induction variables 
  over a subset of the index space of $a$. Elements of $a$ are determined by an
  assignment $a(\bar{e}) = e_r$ where each $e \in
  \bar{e}$ is either a neighbourhood or constant index.
  For all arrays $b \in \bar{b}$ and array subscripts $b(\var{e'})$ in
  $e_r$ and that flow to $e_r$, then each $e' \in \bar{e'}$ is also
  either a neighbourhood or constant index.

  We refer to the assignment $a(\bar{e}) = e_r$ and all the associated
  statements with dataflow to $e_r$ as the \emph{stencil kernel}.
\label{def:stencil}
\end{definition}

Specification declarations associate to a stencil kernel 
a specification for the array access on one or more array
variables $\bar{v}$ within that kernel.

% specifying the shape of indexing expressions for
%arrays $\bar{v}$ that are simultaenously permitted and expected within
%the code to which the specification is given.

\begin{definition}[Consistency, informal]
  A stencil specification defines a \emph{region} associated to an
  array variable $a$ against which a stencil kernel should be
  \emph{consistent}. Informally, the region defined by a specification
  is an $n$-dimensional rectangle crossing the origin.
  Consistency is then a two way requirement.
  On the one hand, all array subscripts contributing to the kernel
  should be consistent with the region; on the other, all parts of the
  region should have an index which fits the space. 
\end{definition}
\todo{Improve the above- haven't quite introduce enough yet to say
more}

\begin{figure}[t]
\vspace{-0.5em}
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{N}_{>0}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
                        \texttt{::} \; v \\
\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[0.4em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{spec} ::= & [\nonterm{mult},] \; [\nonterm{approx},] \; \nonterm{region} \\
\nonterm{mult} ::= & \term{readOnce} \\
\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \nonterm{rvar} \; \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenRefl{\dimTy}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointwise}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointwise}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointwise}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \, \term{+}
  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.5em]
%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}  
%\\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-0.5em]
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\vspace{-0.8em}
\end{figure}

@@

A \term{forward} region of depth $n$ specifies continguous
neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
inclusively for some induction variable $i$.  Similarly, a
\term{backward} region of depth $n$ specifies contiguous neighbour
indices in dimension $d$ from $i$ to $i - n$, and \term{centered}
of depth $n$ from $i - n$ to $i + n$. A \term{pointwise}
stencil specifies a neighbour index $i$. For example the
following shows four specifications with four consistent stencil
kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d} respectively:
%%
\begin{minted}{fortran}
!= stencil forward(depth=2, dim=1) :: a
e(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
!= stencil backward(depth=2, dim=1) :: b
e(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
!= stencil centered(depth=1, dim=2) :: c
e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
!= stencil pointwise(dim=3) :: d
e(i, j) = d(0, 0, i)
\end{minted}
%%
Note that every dimension need not be specified for; in the above, the
specifications on lines $1$ and $5$ and $7$ are the principal specifications
for code on lines $2$ and $6$ and $8$, but the specification on line
$3$ is not principal for line $4$;
we return to the notion of principality later.

The \term{forward}, \term{backward}, and \term{centered} regions may
all have an additional attribute \term{nonpointwise} which marks that
the region does not include indices which are purely base induction
variables, \ie{}, offset by $0$.  For example, the following is a
\term{nonpointwise} \term{backward} stencil
%
\begin{minted}{fortran}
!= stencil backward(depth=2, dim=1, nonpointwise) :: a
b(i) = a(i-1) + a(i-2)
\end{minted}
%
The \term{pointwise} regions cannot have a \term{nonpointwise} attribute.
